### 2025-09-17 paper/abstract.tex
More on abstract.

\begin{abstract}
We present a proof-of-concept implementation and evaluation of an agent-based LLM framework for authoring
\emph{transparent text}, natural language articles which provide interactions allowing a reader to explore the
relationship to the underlying data. We evaluate our approach on a subset of SciGen, a dataset consisting of
tables from scientific articles and their corresponding descriptions.
\end{abstract}

>>>>

\begin{abstract}
We introduce the idea of \emph{transparent text}, web-based data-driven scholarly articles which allow a
reader to explore the relationship to the data set by hovering over fragments of text, and present an
agent-based LLM framework for authoring transparent text. Our approach builds on recent developments in
integrated data provenance for general-purpose programming languages. Our system uses the open source
programming language Fluid, which uses provenance tracking to link visual elements to data, and show how Fluid
can also be used as host platform for transparent text. Our architecture consists of two LLM agents which
support the human author during the creation of a transparent document. A SuggestionAgent identifies fragments
of text which could in principle be computed from data, including numerical values selected from records or
computed by aggregations like sum and mean; comparatives and superlatives like ``better than'' and
``largest''; trend-adjectives like ``growing''; and other idiomatic quantitative or semi-quantitative phrases.
An InterpretationAgent synthesises suitable Fluid queries over the data. We evaluate our approach on a subset
of SciGen, a dataset consisting of tables from scientific articles and their corresponding descriptions, which
we extend with hand-generated counterfactual test cases for evaluating how machine-generated expressions
generalise in the presence of changes to the underlying data. Our results show that some state-of-the-art
models are able to synthesise compound expressions that generate data-driven text that agrees with gold
solutions.
\end{abstract}
